Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-19 ms.Rnw
SEIRS table

%% \section{Examples}

%% \subsection{SIRS}

%% Was shown in an earlier paper.
%% [Display filter equation and plot a curve.]

%% \subsection{SEIRS}

%% Display filter equation.
%% Discuss the choice of $\pi$ and $q$.
%% Present a likelihood curve.

%% \subsection{Two-strain competition}

%% Display filter equation.
%% Discuss the choice of $\pi$ and $q$.
%% Present a likelihood curve.

%% \subsection{Superspreading model}

%% Display filter equation.
%% Discuss the choice of $\pi$ and $q$.
%% Present a likelihood curve.

%% \subsection{Linear birth-death and Moran models}

%% Exact solution of the filtering equations is possible in the unstructured case, yielding precisely the likelihoods for these two models.
%% This establishes that the present theory is a strict generalization of these cases.
%% It does away with the need for large sample-size and small sample-fraction approximations, the necessity of assuming slow changes in effective population size, and the need for linearization.

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth  & 2 \\
    $\lab{Recov}$  & $\gamma I$        & pure death  & 0 \\
    $\lab{Wane}$   & $\omega R$        & neutral     & 0 \\
    $\lab{Sample}$ & $\psi I$          & pure sample & 1 \\
    \hline\hline
  \end{tabular}
\end{table}

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
The KFE for the population process is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
      &\frac{\beta(t)\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,E,I,R)\\
      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
      -\sigma\,E\,v(t,S,E,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
      -\gamma\,I\,v(t,S,E,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
      -\omega\,R\,v(t,S,E,I,R),
    \end{split}
  \end{equation*}
\end{mathsize}%
where $N=S+E+I+R$ is the total population size.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth     & $(1,1)$ \\
    $\lab{Prog}$   & $\sigma E$        & pure migration & $(0,1)$ \\
    $\lab{Recov}$  & $\gamma I$        & pure death     & $(0,0)$ \\
    $\lab{Wane}$   & $\omega R$        & neutral        & $(0,0)$ \\
    $\lab{Sample}$ & $\psi I$          & pure sample    & $(0,1)$ \\
    \hline\hline
  \end{tabular}
\end{table}

>>>>

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & pure birth  & 2 \\
    $\lab{Recov}$  & $\gamma I$        \bigstrut & pure death  & 0 \\
    $\lab{Wane}$   & $\omega R$        \bigstrut & neutral     & 0 \\
    $\lab{Sample}$ & $\psi I$          \bigstrut & pure sample & 1 \\
    \hline\hline
  \end{tabular}
\end{table}

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
The KFE for the population process is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
      &\frac{\beta\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
      -\frac{\beta\,S\,I}{N}\,v(t,S,E,I,R)\\
      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
      -\sigma\,E\,v(t,S,E,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
      -\gamma\,I\,v(t,S,E,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
      -\omega\,R\,v(t,S,E,I,R),
    \end{split}
  \end{equation*}
\end{mathsize}%
where $N=S+E+I+R$ is the total population size.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

\input{figs/seirs_events}

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & pure birth     & $(1,1)$ \\
    $\lab{Prog}$   & $\sigma E$        \bigstrut & pure migration & $(0,1)$ \\
    $\lab{Recov}$  & $\gamma I$        \bigstrut & pure death     & $(0,0)$ \\
    $\lab{Wane}$   & $\omega R$        \bigstrut & neutral        & $(0,0)$ \\
    $\lab{Sample}$ & $\psi I$          \bigstrut & pure sample    & $(0,1)$ \\
    \hline\hline
  \end{tabular}
\end{table}

\begin{table}[h!]
  \caption{
    \label{tab:seirs_integ}
    Elements of a scheme for computing the likelihood under the SEIRS model.
    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{c c c c c c c | c}
    \hline\hline
    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & $Q$ & Line\\[2ex]
    \hline
    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I-\ell_I}{I}$ & $1$ & 1 \\
    \cline{3-8}
    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2 I}$ & $1$ & 2 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2 I}$ & $1$ & 3 \\
    \cline{3-8}
    && $(1,1)$ & --- & --- & 0 & $0$ & 4 \\
    \cline{2-8}
    &\multirow{2}{*}{$\lab{Prog}$}
    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E-\ell_E}{E}\,\Indicator{E>\ell_E}$ \bigstrut & $1$ & 5 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E}\,\Indicator{E>\ell_E}$ \bigstrut & $1$ & 6 \\
    \cline{2-8}
    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & $1$ & 7 \\
    \cline{2-8}
    & $\lab{Wane}$ & $(0,0)$ & $1$ \bigstrut & \unbroken \bigstrut & $1$ & $1$ & 8 \\
    \cline{2-8}
    & $\lab{Sample}$ & $\ast$ & --- & --- & 0 & 0 & 9 \\
    \hline
    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & \multirow{4}{*}{$(1,1)$}
    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & $1$ & 10 \\
    \cline{4-8}
    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & $1$ & 11 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & $0$ & 12 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & $0$ & 13 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & $0$ & 14 \\
    \hline
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,0)$}
    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & $1$ & $1$ & 15 \\
    \cline{4-8}
    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 16 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & $0$ & 19 \\
    \hline
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,1)$}
    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & $1$ & $1$ & 17 \\
    \cline{4-8}
    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 18 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & $0$ & 19 \\
    \hline\hline
  \end{tabular}
\end{table}

A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
Given a particular choice of importance-sampling kernel, the filtering equation follows.
For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R)=
      &\frac{\beta\,(S+1)\,I}{N}\,w(t,S+1,E-1,I,R)
      -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R)\\
      &+\sigma\,(E+1)\,w(t,S,E+1,I-1,R)
      -\sigma\,E\,w(t,S,E,I,R)\\
      &+\gamma\,(I+1)\,w(t,S,E,I+1,R-1)
      -\gamma\,I\,w(t,S,E,I,R)\\
      &+\omega\,(R+1)\,w(t,S-1,E,I,R+1)
      -\omega\,R\,w(t,S,E,I,R),
    \end{split}
  \end{equation*}
\end{mathsize}%
